Testing Charts: viewer's perceptual accuracy in surveys

### 2023-07-18 SDSS-2023/testing-charts-accuracy.qmd
adding introduction text and doing some restructuring of the middle sections

Outline for introduction:

-   What do viewers see when we show them a data chart?

-   Visuals allow us to make comparisons between values

-   Design of a data visualization impacts viewer ability to make those comparisons

    -   And more broadly, impacts how viewers interact with a chart

-   There is prior work studying this topic

    -   Cleveland and McGill laid the foundations

    -   Mackinlay expanded to theoretical scales

    -   Heer/Bostock expanded to larger study population

-   However, prior work has utilized specific populations of study respondents.
    These populations are not representative of the general population, a common target audience for data visualization and scientific communication work

-   Our work focuses on the following:

    -   Utilizing a survey panel as a mechanism to test perception in data visualizations

    -   Testing whether prior results from convenience samples hold with a nationally representative sample

    -   Studying variations in data visualization design and their impact on respondent behavior and accuracy of participants' responses.

    -   \kr{Include somewhere in here:} The advantage of using a probability-based approach is two-fold.
        First, we have access to a large sample of survey participants and thus have greater power in making inference about graphical perceptional abilities.
        Second, the sample is representative of the general adult public in the U.S., which is an important target audience for scientific communication.

-   This paper presents a series of tests completed and the resulting findings.

    -   We consider structural (and aesthetic?) variations on bar charts and pie charts

    -   Our tests are centered around the following central research questions:

        -   How do structural design choices in a data visualization impact viewers' ability to identify the larger of two elements?

        -   How is viewer behavior (zooming, time spent on question, certainty of response) impacted by structural and aesthetic design choices in a data visualization?

-   The remainder of the paper is organized as follows:

    -   we first describe the stimulus design

    -   then we describe the study population (and survey design?)

    -   subsequently, we share analyses of resulting survey responses across a series of stimuli

    -   finally, we discuss implications of this study and next steps.

What do viewers see, when we show them a data chart?
A crucial step in the process of 'understanding' a chart, is that a visual allows us to make comparisons between its parts.
@clevelandGraphicalPerceptionTheory1984 defined for the purpose of their seminal study the better visual as the one that allows viewers to make more accurate comparisons.
Based on mappings of quantitative variables to different graphical elements, Cleveland and McGill's study resulted in a ranking of perceptual tasks from most accurate to least accurate.
@mackinlayAutomatingDesignGraphical1986 took this approach and expanded the tasks and their order to ordinal and nominal scales, as shown in @fig-perceptual-tasks.

@@

What we need to remember, is that Mackinlay's work is purely theoretical.
Cleveland and McGill's rankings based on a study.
However, this study is a small convenience sample, consisting of only a few individuals recruited from among the authors' coworkers and their spouses.
@heerCrowdsourcingGraphicalPerception2010 have reproduced Cleveland and McGill's rankings using a crowd sourcing platform.
A total of XXX amazon turkers were involved.

Crowd workers are known to be biased towards male, young and relatively higher education: @borgoCrowdsourcingInformationVisualization2017

Here, we seek to -- first -- answer the question whether it is possible to use a survey to reproduce (some of) the rankings.

@@

**Why do we utilize just-noticeable differences?** We employ comparisons at the JND in our tasks in order to maximize our ability to identify the impact of design changes on viewer accuracy and behavior.
Asking perception tasks in a survey differs from the controlled environment of a cognitive lab, where these kind of questions may usually be assessed.
Rather than asking the same (or similar) type of question with varied signal strength multiple hundred times, we are limited to only a few questions at a time.
With a small set of tasks, we need to present tasks that are perceptually hard, and thus ask questions about stimuli that are close to our perceptual threshold.
Therefore, we focus on questions which vary the presented image, but ask viewers to compare the same values across those varied images.

@@

**We vary the task by presenting viewers with different structural representations of the same data.**

-   Stimuli within context of data viz

    -   Title, legend, axes, etc.

-   Results are organized as follows:

    -   .... Haven't decided yet.

-   How do structural design choices in a data visualization impact viewers' ability to identify the larger of two elements?

-   How do aesthetic design choices in a data visualization impact viewers' ability to identify the larger of two elements?

-   How is viewer behavior (zooming, time spent on question, certainty of response) impacted by structural and aesthetic design choices in a data visualization?

Structural design choices

Mapping, Stacked bar, Vertical, Horizontal, Horizontal wide

Facetted bar

Only have a split sample for this

Pie, Alignment

We have this for all above mappings, but the setup is a little different for facetted bar

Aesthetic design choices (structural choices seems stronger/there could be a lot to talk about there... should we skip aesthetic on this one?)

Colors

Use of gridlines

Outcomes/responses for modeling:

Binary accuracy (correct/incorrect -- 'they are the same' is incorrect here)

Ordinal response (a/b/they are the same)

Time spent on question (continuous, in seconds)

Certainty

I've noted this below, but: how to model?
Ordinal response?
Binary (certain or very certain vs everybody else)?

## Survey setup - Stimulus description

\autoref{fig-tasks-12} shows the two stacked barcharts shown to participants in the survey.
The marked tiles in each plot are 155 pixels apart.
Based on @luModelingJustNoticeable2022's model, a difference of 155 pixels leads to a just noticeable difference of 3.5 pixels.
The heights of the bars are 205 (left) and 213 pixels (right), respectively, corresponding to about twice the JND.
This difference should lead to a relatively high accuracy rate for participants and simultaneously limit the amount of frustration resulting from a task that is perceived as 'too hard'.

@@

DATA THAT MAY BE INCLUDED IN ANALYSIS:

ROUNDS 1-2: Color variations on vertical stacked bar, aligned and unaligned

ROUND 3: Horizontal and horizontal wide, aligned and unaligned

ROUND 6: Facetted bar (split sample w/o forcing choice)

ROUND 7: Aligned vs unaligned pie (full sample)

@@

Test 1

@@

Test 2

-   Structure of stacked bars
    -   Accuracy and responses

@@

Test 3

@@

How should we group results?

-   By stimulus/structural variation?

-   By response (binary, ordinal, certainty, time spent, etc.)

-   By accuracy vs by behavior (zooming, time spent, etc.)

>>>>

What do viewers see when we show them a data chart?
A data chart -- at its core -- maps quantitative values to graphical elements representing their relative values.
Modern data visualizations are much more than a simple, objective mapping of values to a plane; they contain contextual and design elements, and are often structured to support the viewer in understanding a particular view of a set of data or specific pattern underlying the values.
The design of a data visualization impacts a viewer's ability to achieve that understanding; a poorly designed data visualization may leave viewers struggling to understand the content or context, or make it difficult to complete accurate and useful comparison of values across groups or time points.
More broadly, the design of a data visualization can change how viewers interact with the chart.

A crucial step in the process of interacting with and understanding a chart is the viewer's employment of comparisons of the parts within.
@clevelandGraphicalPerceptionTheory1984 observed as such, and in their seminal study defined the better visual among a pair as the one that allows viewers to make more accurate comparisons.
Based on mappings of quantitative variables to different graphical elements, Cleveland and McGill's study resulted in a ranking of perceptual tasks from most accurate to least accurate, which was then extended by @mackinlayAutomatingDesignGraphical1986 to a theoretical framework ranking tasks' order along their ordinal and nominal scales, as shown in @fig-perceptual-tasks.

Cleveland and McGill's work -- while a foundational user study in graphical perception -- utilized a small convenience sample, consisting of only a few individuals recruited from among the authors' coworkers and their spouses.
@heerCrowdsourcingGraphicalPerception2010 reproduced Cleveland and McGill's rankings using a larger sample from a crowd sourcing platform.
A total of XXX Amazon Mechanical Turkers were involved.
Crowd sourced samples were shown by @borgoCrowdsourcingInformationVisualization2017 to be biased towards more male, younger, and relatively higher education relative to the adult U.S. population as a whole.
These study populations are thus not representative of the general population, a common target audience for data visualization and scientific communication work.
Further, the populations' emphasis on higher education individuals also leads to results which hold for groups of individuals who may be more likely to have prior exposure to data visualization in the context of scientific communication, or more exposure to data topics in higher education, but may not hold across other groups within the population.
\kr{I don't know if the prior statement is going too far? Maybe we can rephrase.}

\kr{The mackinlay thing feels like maybe we are overemphasizing the ranking done by cleveland/mcgill and then mackinlay - I think we could emphasize less if we want}

Our work -- as that of Cleveland and McGill and Heer and Bostock -- centers around studying data visualization design choices and their impact on viewer behavior and accuracy of viewers' responses.
We seek to answer whether it is possible to reproduce some of their findings in the context of a survey with a large, nationally-representative set of respondents, and within that context we focus on the following research questions:

1.  How do structural design choices in a data visualization impact viewers' ability to identify the larger of two elements?

2.  How is viewer interaction with the task impacted by structural design choices in a data visualization?

We employ a probability-based survey panel and run a series of perception tests with nationally-representative samples of respondents from that panel.
The advantage of using a probability-based approach is two-fold.
First, we have access to a large sample of survey participants and thus have greater power in making inference about graphical perceptional abilities.
Second, the sample is representative of the general adult public in the U.S., which is an important target audience for scientific communication, and this allows us to test whether prior results from convenience samples hold with a nationally representative sample.
\kr{Do we want a stronger statement about this here?}

We present viewers with structural variations on bar charts and pie charts and ask them to answer questions comparing the size of elements within those charts.
In this work, we present the series of tests we completed and the resulting findings.
The remainder of the paper is organized as follows: first, we describe the design of the visual stimulus used in our perception tests.
We then describe the population of study respondents and obtained survey sample.
Subsequently, we share our analyses of the resulting survey responses across each of our tests, including analyses on accuracy of responses and response behavior.
Finally, we discuss implications of this work and next steps.

@@

\kr{The figure below is still a bit of an orphan, but I think we can revisit the intro ad study setup after we get more of the results in the paper}

@@

**Why do we utilize just-noticeable differences?** We employ comparisons at the JND in our tasks in order to maximize our ability to identify the impact of design changes on viewer accuracy and behavior.
Asking perception tasks in a survey differs from the controlled environment of a cognitive lab, where these kind of questions may usually be assessed.
Rather than asking the same (or similar) type of question with varied signal strength dozens or hundreds of times, we are limited to only a few questions at a time.
With a small set of tasks, we need to present tasks that are perceptually hard, and thus ask questions about stimuli that are close to our perceptual threshold.
Therefore, we focus on questions which vary the presented image, but ask viewers to compare the same values across those varied images.

@@

\kr{Somewhere here we should also talk about how we measure viewer behavior/interaction with the chart: certainty, time spent, zooming, devices, etc. }

### Test 1: Alignment in stacked bar charts

\autoref{fig-tasks-12} shows the two stacked bar charts shown to participants in the first test.
The marked tiles in each plot are 155 pixels apart.
Based on @luModelingJustNoticeable2022's model, a difference of 155 pixels leads to a just noticeable difference of 3.5 pixels.
The heights of the bars are 205 (left) and 213 pixels (right), respectively, corresponding to about twice the JND.
This difference should lead to a relatively high accuracy rate for participants and simultaneously limit the amount of frustration resulting from a task that is perceived as 'too hard'.

@@

### Test 2: Orientation of stacked bar charts

Let's talk about the stacked bar flippy setup here

### Test 3: Alignment and orientation in pie charts

Let's talk about the pies here

@@

### Test 1

@@

### Test 2

-   Structure of stacked bars -- vertical, horizontal, horizontal wide
    -   Accuracy and responses

@@

### Test 3

@@
